---
title: "Niche climate exposure workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Niche climate exposure workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`climniche` assesses niche climate exposure: projected climate change
interpreted relative to the climate conditions currently associated with a
species. The workflow asks whether future climate at current occurrence,
range, or thresholded SDM cells becomes closer to the current realised climatic
niche, farther from it, or beyond an empirical niche boundary.

The package is designed for researchers who have current climate data, future
climate projections and occurrence records, a range map, or an SDM suitability
map. It does not require demographic, dispersal or physiological data.

```{r}
library(climniche)
```

## 1. Fit niche climate exposure

This simulated example creates two future scenarios with the same climate
change amount but opposite niche meanings.

```{r}
sim <- simulate_climniche(n = 1000, p = 3, rho = 0.5, shift = 0.5)

divergence <- fit_climniche(
  current = sim$current,
  future = sim$future_away,
  occupied = sim$occupied,
  center = sim$center,
  sensitivity = sim$sensitivity,
  scale = FALSE
)

convergence <- fit_climniche(
  current = sim$current,
  future = sim$future_toward,
  occupied = sim$occupied,
  center = sim$center,
  sensitivity = sim$sensitivity,
  scale = FALSE
)
```

## 2. Read the analysis as a report

`climniche_report()` returns a plain language summary for ecological
interpretation.

```{r}
rep <- climniche_report(divergence, species = "simulated species")
rep
```

The complete table for individual cells can be extracted for further analysis.

```{r}
head(climniche_table(divergence))
```

For a saved record, write a Markdown report.

```{r, eval = FALSE}
write_climniche_report(rep, "climniche_report.md")
```

## 3. Visualise the results

The exposure plane is a quick diagnostic. The x axis shows how much climate
changes. The y axis shows whether future climate becomes farther from or closer
to the current realised climatic niche.

```{r, fig.width = 6.2, fig.height = 4.2}
plot_climniche_exposure(divergence)
```

The niche climate diagram gives the same calculation a spatial interpretation
inside species weighted climate space. The grey polygon is the current realised
niche boundary, the green outline is the future climate area and the arrows
show mean change by exposure class.

```{r, fig.width = 6.2, fig.height = 4.8}
plot_climniche_diagram(divergence)
```

For a manuscript-style summary, use `plot_climniche_showcase()`. This produces
the binned exposure plane, class proportions, variable contributions and metric
distributions from the same fitted object.

```{r, fig.width = 7.2, fig.height = 5.2}
plot_climniche_showcase(divergence)
```

Class proportions provide a compact summary for non specialist readers.

```{r, fig.width = 6.2, fig.height = 3.4}
plot_climniche_class_summary(divergence)
```

Variable contributions indicate which climate variables make future climate
less similar or more similar to current niche conditions.

```{r, fig.width = 5.6, fig.height = 3.4}
plot_climniche_variable_contribution(divergence)
```

## 4. Raster workflow

For spatial analyses, use `fit_climniche_raster()` or `fit_climniche_terra()`
with current and future climate rasters and a presence, range, or SDM
suitability raster. Continuous suitability rasters are thresholded: cells with
values greater than `occupied_threshold` define the current distribution used
to estimate the realised niche centre and boundary.

```r
fit <- fit_climniche_raster(
  current = clim_current,
  future = clim_future,
  occupied = range_raster,
  center = niche_center,
  sensitivity = variable_sensitivity,
  boundary = 0.95
)

plot_climniche_map(fit, metric = "niche_distance_change",
                   occupied = range_raster)
plot_climniche_map(fit, metric = "outside_niche_exceedance",
                   occupied = range_raster)
plot_climniche_classes(fit, occupied = range_raster)
```

## 5. Interpretation boundaries

Niche climate exposure is an exposure interpretation. It does not estimate
persistence, abundance change, dispersal limitation, adaptive capacity or
conservation priority by itself. It is most useful as a diagnostic layer that
can be combined with SDMs, trait data, demographic models or expert knowledge
when those data are available.
