---
title: "Niche relative geometry"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Niche relative geometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```

## Estimand

`climniche` describes projected climatic change at a site relative to a fixed
current niche reference. Let $c_i$ and $f_i$ be current and future climate,
$\mu$ the weighted centre of the current reference, and $d_A$ the distance
under positive semidefinite climatic weighting matrix $A$:

$$
d_A(x,y)=\sqrt{(x-y)^\mathsf{T}A(x-y)}.
$$

$$
\begin{aligned}
D_i &= d_A(f_i,c_i),\\
R_i &= d_A(f_i,\mu)-d_A(c_i,\mu),\\
C_i &= \sqrt{\max(0,D_i^2-R_i^2)},\\
E_i &= \max(0,d_A(f_i,\mu)-B_q).
\end{aligned}
$$

Climatic Displacement ($D_i$) measures local current-to-future change. Niche
Distance Shift ($R_i$) records whether future conditions lie closer to or
farther from the reference centre. Climatic Reconfiguration ($C_i$) is the
derived non-radial component. If $r_{0i}$ and $r_{1i}$ are current and future
distances from the centre and, when both are positive, $\theta_i$ is the angle
between their centred vectors, then

$$
C_i^2=2r_{0i}r_{1i}\{1-\cos(\theta_i)\}.
$$

This identity does not make $C_i$ a strict orthogonal projection or an
independent ecological process. Niche Boundary Exceedance ($E_i$) is a
separate future-state comparison with the weighted empirical radial boundary
$B_q$.

## Controlled paths

Five paths isolate the terms in two climatic dimensions. Each site starts at
$(1,0)$, one unit from a fixed centre at $(0,0)$.

```{r}
library(climniche)

current <- matrix(
  rep(c(1, 0), 5),
  ncol = 2,
  byrow = TRUE,
  dimnames = list(
    c("unchanged", "toward", "away", "reoriented", "crossed centre"),
    c("climate 1", "climate 2")
  )
)
future <- rbind(
  unchanged = c(1, 0),
  toward = c(0, 0),
  away = c(2, 0),
  reoriented = c(0, 1),
  `crossed centre` = c(-1, 0)
)
colnames(future) <- colnames(current)

geometry_fit <- fit_climniche(
  current = current,
  future = future,
  occupied = rep(1, nrow(current)),
  center = c(0, 0),
  A = diag(2),
  scale = FALSE,
  preprocess = FALSE,
  boundary = 0.95,
  tolerance = 0
)
```

```{r}
geometry_table <- data.frame(
  path = rownames(current),
  `Climatic Displacement` = geometry_fit[["climate_change_amount"]],
  `Niche Distance Shift` = geometry_fit[["niche_distance_change"]],
  `Climatic Reconfiguration` =
    geometry_fit[["climate_reconfiguration"]],
  `Niche Boundary Exceedance` =
    geometry_fit[["niche_boundary_exceedance"]],
  check.names = FALSE
)

knitr::kable(geometry_table, digits = 3)
```

The toward and away paths have equal Climatic Displacement but opposite Niche
Distance Shift. Reorientation changes direction without changing radial
distance. Crossing the centre also changes the direction of the
centre-referenced vector, which is why Climatic Reconfiguration is positive
even in one climatic dimension. Only the away path exceeds the fitted radial
boundary.

```{r}
identity_error <- with(
  geometry_fit,
  climate_change_amount^2 -
    niche_distance_change^2 -
    climate_reconfiguration^2
)
max(abs(identity_error))
```

## Relation to other outputs

The quantities answer a different question from an SDM projection. An SDM
estimates change in a fitted suitability response, whereas `climniche`
describes change in the climatic space used to define the current reference.

| Output | Quantity described | Niche relative direction |
|:--|:--|:--|
| Standardised current-to-future distance | Magnitude of local climatic change | No |
| Climate-niche factor analysis departure | Magnitude in a fitted climate sensitivity space | Not in the original formulation |
| Projected SDM suitability change | Change in the fitted suitability response | Model dependent |
| `climniche` | Local displacement, signed centre-relative shift and empirical radial boundary exceedance | Yes |

Rinnan and Lawler (2019) identified the missing toward-or-away interpretation
as a direction for further development in climate-niche factor analysis.
`climniche` addresses that specific gap at the cell level. The reference centre
and boundary remain empirical summaries of the supplied current distribution;
they are not physiological limits or a fundamental niche.

## Reference

Rinnan DS, Lawler J (2019). Climate-niche factor analysis: a spatial approach
to quantifying species vulnerability to climate change. *Ecography* 42:
1494-1503. [doi:10.1111/ecog.03937](https://doi.org/10.1111/ecog.03937).
