---
title: "Examples"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Examples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.align = "center")
```

This vignette gives short examples for the main data forms accepted by
`climniche`: matrices, binary range rasters and continuous suitability rasters.
The examples use simulated data so that the workflow can run without downloads.
In an applied study, the same objects would usually come from occurrence
records, an SDM, a range polygon, or environmental rasters.

```{r}
library(climniche)
```

## Matrix data

Use the matrix interface when current and future climate values have already
been extracted for the same sites or grid cells. `occupied` can be a logical
vector, row indices, or a numeric vector such as a thresholded SDM prediction.

```{r}
sim <- simulate_climniche(n = 1200, p = 3, rho = 0.45, shift = 0.55)

sdm_score <- as.numeric(sim$occupied)
sdm_score[sdm_score == 1] <- runif(sum(sim$occupied), 0.55, 1.00)
sdm_score[sdm_score == 0] <- runif(sum(!sim$occupied), 0.00, 0.45)

fit_matrix <- fit_climniche(
  current = sim$current,
  future = sim$future_away,
  occupied = sdm_score,
  occupied_threshold = 0.5,
  center = sim$center,
  sensitivity = sim$sensitivity,
  scale = FALSE
)

climniche_summary(fit_matrix)
```

```{r, fig.width = 6.2, fig.height = 4.2}
plot_climniche_exposure(fit_matrix)
```

## Niche climate diagram

The diagram is intended for method explanation and manuscript figures. It
summarises where the current niche lies in climate space and how future climate
moves relative to that niche.

```{r, fig.width = 6.2, fig.height = 4.8}
plot_climniche_diagram(fit_matrix)
```

The same plot can expose the source data when a reviewer or collaborator wants
to inspect the spread of cell-level changes.

```{r, fig.width = 6.2, fig.height = 4.8}
plot_climniche_diagram(fit_matrix, type = "sample", max_arrows = 120)
```

## Compact report figure

`plot_climniche_showcase()` combines four views of the same fit: the exposure
plane, class proportions, variable contributions and metric distributions.

```{r, fig.width = 7.2, fig.height = 5.2}
plot_climniche_showcase(fit_matrix)
```

## Raster data

Raster workflows use the same calculation, but preserve raster outputs for map
making. The example below builds a small artificial climate grid. The
`occupied` raster is continuous, like an SDM suitability map, and cells above
`occupied_threshold` define the current reference niche.

```{r, eval = requireNamespace("raster", quietly = TRUE)}
library(raster)

r1 <- raster(nrows = 30, ncols = 45, xmn = -10, xmx = 35,
             ymn = 30, ymx = 46, crs = "+proj=longlat +datum=WGS84")
xy <- rasterToPoints(r1)[, 1:2]

temp_current <- r1
sal_current <- r1
temp_future <- r1
sal_future <- r1
suitability <- r1

values(temp_current) <- 18 + 0.18 * xy[, 1] - 0.25 * xy[, 2] +
  sin(xy[, 1] / 4)
values(sal_current) <- 37.5 + 0.02 * xy[, 1] + cos(xy[, 2] / 5)
values(temp_future) <- values(temp_current) + 0.8 + 0.04 * xy[, 1]
values(sal_future) <- values(sal_current) + 0.15 * sin(xy[, 1] / 5)

score <- exp(-((xy[, 1] - 15)^2 / 130 + (xy[, 2] - 38)^2 / 24))
values(suitability) <- score

current_stack <- stack(temp_current, sal_current)
future_stack <- stack(temp_future, sal_future)
names(current_stack) <- names(future_stack) <- c("temperature", "salinity")

fit_raster <- fit_climniche_raster(
  current = current_stack,
  future = future_stack,
  occupied = suitability,
  occupied_threshold = 0.35,
  sensitivity = c(temperature = 1.4, salinity = 0.8),
  boundary = 0.95
)
```

```{r, fig.width = 6.4, fig.height = 3.8, eval = requireNamespace("raster", quietly = TRUE)}
plot_climniche_map(
  fit_raster,
  metric = "niche_distance_change",
  occupied = suitability,
  occupied_only = TRUE,
  occupied_threshold = 0.35
)
```

```{r, fig.width = 6.4, fig.height = 3.8, eval = requireNamespace("raster", quietly = TRUE)}
plot_climniche_classes(
  fit_raster,
  occupied = suitability,
  occupied_only = TRUE,
  occupied_threshold = 0.35
)
```

## Terra data

The `terra` interface follows the same argument names. It is useful for larger
rasters and modern spatial workflows.

```{r, eval = requireNamespace("terra", quietly = TRUE)}
library(terra)

current_terra <- rast(current_stack)
future_terra <- rast(future_stack)
suitability_terra <- rast(suitability)

fit_terra <- fit_climniche_terra(
  current = current_terra,
  future = future_terra,
  occupied = suitability_terra,
  occupied_threshold = 0.35,
  sensitivity = c(temperature = 1.4, salinity = 0.8)
)

climniche_summary(fit_terra)
```

```{r, fig.width = 6.4, fig.height = 3.8, eval = requireNamespace("terra", quietly = TRUE)}
plot_climniche_map(
  fit_terra,
  metric = "outside_niche_exceedance",
  occupied = suitability_terra,
  occupied_only = TRUE,
  occupied_threshold = 0.35
)
```

## Report output

Reports are plain Markdown objects that can be printed, written to disk or
combined with a manuscript workflow.

```{r}
report <- climniche_report(fit_matrix, species = "example taxon")
report
```
